---
title: "Health and Economic consequences of Storm and other severe natural calamities in USA"
output: 
 html_document:
  keep_md: yes
---

#Synopsis 

Storms and other natural calamities can cause unprecedented damage to property and pose a serious threat to human life. Therefore, preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

#Data Processing 

The first step is to read the data into a dataframe 
```{r,echo=TRUE,cache=TRUE}
storm <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

Lets check the Event type variable which requires some preprocessing.
```{r,echo=TRUE,cache=TRUE}
length(unique(storm$EVTYPE))
```

```{r,echo=TRUE,cache=TRUE}
et <- tolower(storm$EVTYPE)
et <- gsub("[[:blank:][:punct:]+]", " ", et) # Replace punctuation characters with a space.
length(unique(et))
storm$EVTYPE <- et # Update the event type variable.
```

#Health Impacts due to severe weather events. 
To find the event type which is responsible for the majority of human deaths, the number of fatalities are aggreagted by the event type
```{r,echo=TRUE,cache=TRUE}
library(plyr)
deaths_injuries <- ddply(storm,.(EVTYPE),summarize,fatalities=sum(FATALITIES),injuries=sum(INJURIES))
```

Find the top 15 events that caused the most fatalities and the most injuries
```{r,echo=TRUE,cache=TRUE}
fatal_events <- head(deaths_injuries[order(deaths_injuries$fatalities,decreasing=TRUE),],15)
injury_events <- head(deaths_injuries[order(deaths_injuries$injuries,decreasing=TRUE),],15)
```

Top 15 events that caused the majority of deaths :-
```{r,echo=TRUE,cache=TRUE}
fatal_events[,c("EVTYPE","fatalities")]
```

Top 15 events that caused the majority of injuries :-
```{r,echo=TRUE,cache=TRUE}
injury_events[,c("EVTYPE","injuries")]
```

#Economic Impacts due to severe weather events.
To analyze the economic impacts on the property and crops during severe event types first we need to convert the the crop and property damage exponent into equivalent numeric value.
```{r,echo=TRUE,cache=TRUE}
exp_convert <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}

p <- sapply(storm$PROPDMGEXP,FUN=exp_convert)
storm$prop_dmg <- storm$PROPDMG * (10 ** p)
c <- sapply(storm$CROPDMGEXP, FUN=exp_convert)
storm$crop_dmg <- storm$CROPDMG * (10 ** c)
```

Compute the economic losses by event type.
```{r,echo=TRUE,cache=TRUE}
library(plyr)
eloss <- ddply(storm, .(EVTYPE),summarize,prop_dmg=sum(prop_dmg),crop_dmg=sum(crop_dmg))
eloss <- eloss[(eloss$prop_dmg >0 | eloss$crop_dmg > 0),]
prop_dmg_events <- head(eloss[order(eloss$prop_dmg,decreasing=TRUE),],15)
crop_dmg_events <- head(eloss[order(eloss$crop_dmg,decreasing=TRUE),],15)
```

Top 15 events that caused majority of property damages :-
```{r,echo=TRUE,cache=TRUE}
prop_dmg_events[,c("EVTYPE","prop_dmg")]
```

Top 15 events that caused majority of crop damages :-
```{r,echo=TRUE,cache=TRUE}
crop_dmg_events[,c("EVTYPE","crop_dmg")]
```

#Results 

The following plot shows the top 15 events that caused the most fatalities and injuries due to severe weather events.
```{r,echo=TRUE}
library(ggplot2)
library(gridExtra)
g1 <- ggplot(data=fatal_events,aes(x=reorder(EVTYPE,fatalities),y=fatalities))+geom_bar(stat="identity",fill="firebrick1",width=0.8)+theme_bw()+labs(x="Event Type",y="Total Fatalities",title="Total fatalities by different events")+theme(legend.position="none")+coord_flip()

g2 <- ggplot(data=injury_events,aes(x=reorder(EVTYPE,injuries),y=injuries))+geom_bar(stat="identity",fill="blueviolet",width=0.8)+theme_bw()+labs(x="Event Type",y="Total Injuries",title="Total injuries by different events")+theme(legend.position="none")+coord_flip()
grid.arrange(g1,g2)
```

Thus, from these plots it is clear that tornados are the main reason for the most number of fatalities and injuries.

The following plot shows the top 15 events that caused the most property damages due to severe weather events.
```{r,echo=TRUE}
library(ggplot2)
library(gridExtra)
g3 <- ggplot(data=prop_dmg_events,aes(x=reorder(EVTYPE,prop_dmg),y=prop_dmg))+geom_bar(stat="identity",fill="springgreen",width=0.8)+theme_bw()+labs(x="Event Type",y="Total Property damages",title="Total property damage by different events")+theme(legend.position="none")+coord_flip()

g4 <- ggplot(data=crop_dmg_events,aes(x=reorder(EVTYPE,crop_dmg),y=crop_dmg))+geom_bar(stat="identity",fill="slategray",width=0.8)+theme_bw()+labs(x="Event Type",y="Total Crop damages",title="Total crop damage by different events")+theme(legend.position="none")+coord_flip()
grid.arrange(g3,g4)
```

From the plot it is clear that flash flood causes most of the property damages and that drought causes most of the crop damages.